Delocalization of Eigenvectors in Random Matrices

### 2025-03-22 Draft.tex
up to diff'

\begin{boxt}{Conditions}
\begin{enumerate}[label=(8.\arabic*)]
\item $\mathbb{E}\left[H_{ij}^\ell \left(\overline{H}_{ij}\right)^{k-\ell}\right] = \mathbb{E}\left[\left(H_{ij}^G\right)^\ell \left(\overline{H}_{ij}^G\right)^{k-\ell}\right], \quad 0\leq \ell\leq k\leq 5$
\item $\forall q\geq 1, \quad \mathbb{E}\left\vert H_{xy}\right\vert ^{2q}\prec W^{-q}$
\end{enumerate}
\end{boxt}
The first condition $(8.1)$ is the $\textit{Five Moment}$ matching assumption that gives the name of our result. The second $(8.2)$ is a moment decay condition that allows us to relax the Gaussian assumption, while still providing sufficient strength of our bounds. However, we still have to verify that each consecutive replacement is manageable. For this purpose, we will define the interpolation for $t\in [0, 1]$ at entries $(ij)$ and $(ji)$, where with slight abuse of notation, we let:
$$tH_{ij} = H^{(i, j)}[H_{ij}, H_{ji}\mapsto t H_{ij}, tH_{ji}]$$
$$tH_{ij}^{G} = H^{(i, j),G}[H^G_{ij}, H^G_{ij}\mapsto t H^G_{ij}, t H^{G}_{ij}].$$
Whenever the replacement step $(i, j)$ is implicit from context, we will drop the superscript. In order to make our theorem applicable to the context of delocalization, we want to consider the following function: 
$$F(H) = \left\vert
(H-z)_{ij}^{-1} - m(z)S_{ij}
\right\vert ^{2p} = \left\vert 
G_{ij}(z)-m(z)S_{ij}
\right\vert ^{2p} = |W_{ij}|^{2p}$$
Our reasoning is simple. Having familiarized ourselves with the flow method and the resolvent approach, it makes sense for us to choose a function that quantifies the deviation of resolvent entries form their deterministic approximations. By raising to power $2p$, we have better tools at our disposal, namely Chebyshev' and Markov's inequalities. Let us apply our strategy to $F$ at step $ij$ for $t\in [0, 1]$:

@@

\noindent By the chain rule for any $\ell$ and $m$:
$$\frac{d}{dt}\left[\partial_{H_{ij}}^\ell \partial_{H_{ji}}^m F(tH_{ij})\right] = \partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^m F(tH_{ij})H_{ij} + \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{m+1} F(tH_{ij})\overline{H_{ij}}$$
Substituting in the derivative formula gives us:
$$
= \sum_{\ell=0}^k \binom{k}{\ell}\left[\partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^{k-\ell} F(tH_{ij})H_{ij} + \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})\overline{H_{ij}}\right]H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell}$$
$$
= \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^{k-\ell} F(tH_{ij})H_{ij}^{\ell+1} \left(\overline{H_{ij}}\right)^{k-\ell} + \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell+1}
$$
$$
= \sum_{\ell=1}^{k+1} \binom{k}{\ell-1}\partial_{H_{ij}}^{\ell} \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^{k+1-\ell} \left(\overline{H_{ij}}\right)^{k+1-\ell} + \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell+1}
$$
But observe that we can split the sums and group it as follows:
$$=\binom{k}{0}\partial_{H_{ij}}^0\partial_{H_{ji}}^{k+1} F(tH_{ij})\left(\overline{H_{ij}}\right)^{k+1}+\binom{k}{k}\partial_{H_{ij}}^{k+1}\partial_{H_{ji}}^{0} F(tH_{ij})\left(\overline{H_{ij}}\right)^{k+1}+$$
$$=+\sum_{\ell=1}^k \left(\binom{k}{\ell-1}+\binom{k}{\ell}\right)\partial_{H_{ij}}^{\ell} \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^{\ell} \left(\overline{H_{ij}}\right)^{k+1-\ell} $$
But since $\binom{k}{k} = \binom{k+1}{k+1}$ and $\binom{k}{\ell-1}+\binom{k}{\ell} = \binom{k+1}{\ell}$, this completes the induction:
$$\frac{d^{k+1}}{dt^{k+1}} F(tH_{ij}) = \sum_{\ell=0}^{k+1} \binom{k+1}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k+1-\ell}$$

\subsection*{Taylor series}

\noindent By Taylor expansion around $t=0$: 
$$F(H_{ij}) =F(0) + \sum_{k=1}^5 \frac{1}{k!} \frac{d^k}{dt^k}\left. F(tH_{ij})\right\vert_{t=0?1}+ \frac{1}{720} \frac{d^6}{dt^6}\left.F(tH_{ij})\right\vert_{t=t_{ij}},$$
where by Taylor's theorem? $t_{ij}\in [0, 1]$ reflects all of the remaining correlations. Similarly,

$$F(H_{ij}^G) =F(0) + \sum_{k=1}^5 \frac{1}{k!} \frac{d^k}{dt^k}\left. F(tH_{ij}^G)\right\vert_{t=0?1}+ \frac{1}{720} \frac{d^6}{dt^6}\left.F(tH_{ij}^G)\right\vert_{t=t_{ij}^G},$$
for another such $t_{ij}^G\in[0, 1]$.
Then, let us take their difference (\textcolor{red}{verify the matrix identity }$\{--\}$:
$$F(H_{ij}) - F(H_{ij}^G) = \sum_{k=1}^5\frac{1}{k!}\sum_{\ell = 0}^k c_{\ell, k} \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell} F(0)\left\{H_{ij}^\ell \left(\overline{H}_{ij} \right)^{k-\ell}- \left(H_{ij}^G\right)^\ell \left(\overline{H}_{ij}^G\right)^{k-\ell}\right\}+$$
$$+\frac{1}{720} \left[\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}-\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}^G}\right]$$
By the moment matching condition $(8.1)$, we have: $$\mathbb{E}\left\vert F(H_{ij} - F(H_{ij}^G)\right\vert\lesssim \mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}\right\vert+\mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}^G}\right\vert$$
Using the formula we found earlier (\textcolor{red}{write out the cancellation by absolute value}):
$$\mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}\right\vert\lesssim \sum_{k=0}^6 |H_{ij}|^k \left\vert \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell} F(tH_{ij})|_{t=t_{ij}} \right\vert$$
Let the LHS and RHS be $\mathcal{P}$ and $\mathrm{Q}$, respectively. We want $\mathcal{Q}\leq W^C$ for some $C \lesssim 1$ , s.t. $\mathcal{Q}\prec 1$ (\textcolor{red}{definition? - revisit}). \\

>>>>

\begin{boxt}{Conditions}
\begin{enumerate}[label=(8.\arabic*)]
\item $\mathbb{E}\left[H_{ij}^\ell \left(\overline{H}_{ij}\right)^{k-\ell}\right] = \mathbb{E}\left[\left(H_{ij}^G\right)^\ell \left(\overline{H}_{ij}^G\right)^{k-\ell}\right], \quad 0\leq \ell\leq k\leq 5$
\item $\forall q\geq 1, \quad \mathbb{E}\left\vert H_{xy}\right\vert ^{2q}\prec W^{-q}$
\end{enumerate}
\end{boxt}
The first condition $(8.1)$ is the $\textit{Five Moment}$ matching assumption that gives the name of our result. The second $(8.2)$ is a moment decay condition that allows us to relax the Gaussian assumption, while still providing sufficient strength of our bounds. However, we still have to verify that each consecutive replacement is manageable. For this purpose, we will define the interpolation for $t\in [0, 1]$ at entries $(ij)$ and $(ji)$, where with slight abuse of notation, we let:
$$tH_{ij} = H^{(i, j)}[H_{ij}, H_{ji}\mapsto t H_{ij}, tH_{ji}]$$
$$tH_{ij}^{G} = H^{(i, j),G}[H^G_{ij}, H^G_{ij}\mapsto t H^G_{ij}, t H^{G}_{ij}].$$
Whenever the replacement step $(i, j)$ is implicit from context, we will drop the superscript. In order to make our theorem applicable to the context of delocalization, we want to consider the following function: 
\begin{equation*}
F(H) = \left\vert
(H-z)_{ij}^{-1} - m(z)S_{ij}
\right\vert ^{2p} = \left\vert 
G_{ij}(z)-m(z)S_{ij}
\right\vert ^{2p} = |W_{ij}|^{2p}\tag{8.3}
\end{equation*}
Our reasoning is simple. Having familiarized ourselves with the flow method and the resolvent approach, it makes sense for us to choose a function that quantifies the deviation of resolvent entries form their deterministic approximations. By raising to power $2p$, we have better tools at our disposal, namely Chebyshev' and Markov's inequalities. Let us apply our strategy to $F$ at step $ij$ for $t\in [0, 1]$:

@@

\noindent By the chain rule for any $\ell$ and $m$:
$$\frac{d}{dt}\left[\partial_{H_{ij}}^\ell \partial_{H_{ji}}^m F(tH_{ij})\right] = \partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^m F(tH_{ij})H_{ij} + \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{m+1} F(tH_{ij})\overline{H_{ij}}$$
Substituting in the derivative formula gives us:
$$
= \sum_{\ell=0}^k \binom{k}{\ell}\left[\partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^{k-\ell} F(tH_{ij})H_{ij} + \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})\overline{H_{ij}}\right]H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell}$$
$$
= \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^{\ell+1} \partial_{H_{ji}}^{k-\ell} F(tH_{ij})H_{ij}^{\ell+1} \left(\overline{H_{ij}}\right)^{k-\ell} + \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell+1}
$$
$$
= \sum_{\ell=1}^{k+1} \binom{k}{\ell-1}\partial_{H_{ij}}^{\ell} \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^{k+1-\ell} \left(\overline{H_{ij}}\right)^{k+1-\ell} + \sum_{\ell=0}^k \binom{k}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell+1} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k-\ell+1}
$$
But observe that we can split the sums and group it as follows:
$$=\binom{k}{0}\partial_{H_{ij}}^0\partial_{H_{ji}}^{k+1} F(tH_{ij})\left(\overline{H_{ij}}\right)^{k+1}+\binom{k}{k}\partial_{H_{ij}}^{k+1}\partial_{H_{ji}}^{0} F(tH_{ij})\left(\overline{H_{ij}}\right)^{k+1}+$$
$$=+\sum_{\ell=1}^k \left(\binom{k}{\ell-1}+\binom{k}{\ell}\right)\partial_{H_{ij}}^{\ell} \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^{\ell} \left(\overline{H_{ij}}\right)^{k+1-\ell} $$
But since $\binom{k}{k} = \binom{k+1}{k+1}$ and $\binom{k}{\ell-1}+\binom{k}{\ell} = \binom{k+1}{\ell}$, this completes the induction:
\begin{equation*}
\frac{d^{k+1}}{dt^{k+1}} F(tH_{ij}) = \sum_{\ell=0}^{k+1} \binom{k+1}{\ell}\partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k+1-\ell} F(tH_{ij})H_{ij}^\ell \left(\overline{H_{ij}}\right)^{k+1-\ell}\tag{8.4}
\end{equation*}

\subsection*{Taylor series}
Having derived the explicit derivatives, we can now continue with the next step in the Lindeberg exchange strategy. By the Lagrange form of Taylor's (quintic) theorem we know $\exists t_{ij}\in [0, 1]$, s.t.:
$$F(H_{ij}) =F(0) + \sum_{k=1}^5 \frac{1}{k!}\left.  \frac{d^k}{dt^k}F(tH_{ij})\right\vert_{t=0}+ \left.\frac{1}{720} \frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}},$$
where $F(sH_{ij})=F(0)$ for $s=0$ represents setting the $H_{ij}$ and $H_{ji}$ to zero. Similarly, $t_{ij}^G\in[0, 1]$, s.t.:
$$F(H_{ij}^G) =F(0) + \sum_{k=1}^5 \frac{1}{k!} \left. \frac{d^k}{dt^k}F(tH_{ij}^G)\right\vert_{t=0}+ \frac{1}{720}\left. \frac{d^6}{dt^6}F(tH_{ij}^G)\right\vert_{t=t_{ij}^G}.$$
Let us take their difference:
\begin{equation*}\begin{array}{l}
F(H_{ij}) - F(H_{ij}^G)=
\end{array}
\end{equation*}
$$ =\sum_{k=1}^5 \frac{1}{k!} \left[\left. \frac{d^k}{dt^k}F(tH_{ij})\right\vert_{t=0}-\left. \frac{d^k}{dt^k}F(tH_{ij}^G)\right\vert_{t=0}\right\vert +\frac{1}{720} \left[\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}-\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}^G}\right]$$
Focusing on the first term, we can u
We can now recall our formula from before, s.t. our first term becomes:
$$\sum_{k=1}^t \left[\right]$$

$$ \sum_{k=1}^5\frac{1}{k!}\sum_{\ell = 0}^k c_{\ell, k} \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell} F(0)\left\{H_{ij}^\ell \left(\overline{H}_{ij} \right)^{k-\ell}- \left(H_{ij}^G\right)^\ell \left(\overline{H}_{ij}^G\right)^{k-\ell}\right\}+$$
$$+\frac{1}{720} \left[\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}-\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}^G}\right]$$
By the moment matching condition $(8.1)$, we have: $$\mathbb{E}\left\vert F(H_{ij} - F(H_{ij}^G)\right\vert\lesssim \mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}\right\vert+\mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}^G}\right\vert$$
Using the formula we found earlier (\textcolor{red}{write out the cancellation by absolute value}):
$$\mathbb{E}\left\vert\left.\frac{d^6}{dt^6}F(tH_{ij})\right\vert_{t=t_{ij}}\right\vert\lesssim \sum_{k=0}^6 |H_{ij}|^k \left\vert \partial_{H_{ij}}^\ell \partial_{H_{ji}}^{k-\ell} F(tH_{ij})|_{t=t_{ij}} \right\vert$$
Let the LHS and RHS be $\mathcal{P}$ and $\mathrm{Q}$, respectively. We want $\mathcal{Q}\leq W^C$ for some $C \lesssim 1$ , s.t. $\mathcal{Q}\prec 1$ (\textcolor{red}{definition? - revisit}). \\
